GEE models

### 2017-10-20 doc/manuscript.Rmd
edits to manuscript as per Tony

Title: Association of non-esterified fatty acid composition with insulin
sensitivity and beta cell dysfunction in the Prospective Metabolism and Islet Cell
Evaluation (PROMISE) cohort

@@

Word count - Main text: 3869 / 4000 (with figure captions)

<!-- Table and figure count: 4 / 4 -->

<!-- Reference limit 49 / 50 -->

@@

Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on the pathogenesis of diabetes, specifically on differences in the insulin sensitivity and beta cell function over 6 years in a cohort of individuals who are at-risk for diabetes.

@@

At each examination, an 8-12 hour fasting blood sample was drawn from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
Modular's enzymatic colorimetric tests (Roche Canada, Mississauga, Ontario, Canada).  Both insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyser and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form (Linco Res. Inc, St. Charles, Missouri, USA), and has a coefficient of variation (CV) of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics, Mississauga, Ontario, Canada) with a detection range of 0.11 (2 mg/dl) to 41.6 mmol/l.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorised using the 2006 WHO criteria [@WHO2006].

@@

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design an
auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE
models, though other matrices (eg. exchangeable) had similar fit (data not
shown). GEE is well suited to longitudinal cohort studies given it's capacity to
handle any missed visits and for the correlation of variables between time points. 
See the ESM Methods for more detail on GEE.
The predictor variables and continuous covariates were
scaled (mean centered and standardised). The NEFA variables were classified as 
*time-independent* (held constant) as they were measured only at the baseline 
visit, while the outcome variables were set as *time-dependent*. No imputation
was conducted on missing values and incident diabetes cases were excluded from 
the analysis (follow-up data were included until date of diagnosis). A GEE model
is applied for each combination of outcome and NEFA (as either mol% or nmol/ml).

@@

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a]
and the package pls `r packageVersion('pls')` for the PLS analysis.
The R code and output for all analyses of this manuscript is available at 
https://dx.doi.org/10.6084/m9.figshare.3472433. Results were considered
statistically significance at p<0.05, after adjusting for multiple testing using
the Benjamini-Hochberg False Discovery Rate (FDR) [@Benjamini1995a]. STROBE was used
as a guideline for reporting [@Vandenbroucke2007b].
The cohort datasets analysed during the current study are available from the
corresponding author on reasonable request.

@@

sig_fa_unadj <- gee_results %>% 
    dplyr::filter(model == "Unadjusted", p.value < 0.05) %>% 
    dplyr::select(Yterms, Xterms, unit) %>% 
    dplyr::mutate(Xterms = ifelse(Xterms == "Total", "total NEFA", as.character(Xterms)))
sig_fa_adj <- gee_results %>% 
    dplyr::filter(model == "Adjusted", p.value < 0.05) %>% 
    dplyr::select(Yterms, Xterms, unit) %>% 
    dplyr::mutate(Xterms = ifelse(Xterms == "Total", "total NEFA", as.character(Xterms)))
```

@@

For this study, there were `r female$npct` females and `r european$npct` had
European-ancestry (see `r cite_t("basic")`). The mean follow-up time was 
`r calc_followup_time()` years, where `r calc_pct_full_visits()`% of 
participants attended all three visits.
The average age of the participants at baseline was `r aide::average(ds0yr$Age)`
years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` kg/m^2^ (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r famhist$npct`, had a family history of diabetes. Insulin
sensitivity and beta cell function measures showed a significant median
decline from baseline to 6-years of between `r chg_outcomes$chg` 
(p`r chg_outcomes$p`, based on a time-adjusted GEE model) in this analysis.
Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
and `r calc_incid_dysgly("ConvertPreDM")` incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
cases were excluded from GEE analyses (n={{num}} at 3rd year, n=
{{num}} at 6th year). Follow-up data were included until date of
diagnosis.
Certain medications (e.g. lipid medications such as statins or
fibrates) could influence the association between NEFA and the outcomes.
There were `r meds_use[1]` and `r meds_use[2]` participants who used lipid and
cholesterol modifying medications at the baseline and 6-year visit,
respectively. Use of these medications did not influence NEFA composition, the
outcomes, or the GEE models (data not shown).

@@

We used indices of insulin sensitivity and beta cell function estimated from
glucose and insulin values and as such could that lead to measurement bias.
While these indices have all been validated against gold standard measures,
it is possible our measures are not sensitivity enough to identify potential
associations with NEFA.
Finally, our cohort consists of individuals at-risk for diabetes,
who are primarily female and European-ancestry and, while sex and ethnicity did
not influence the results presented from this cohort, our findings may not
entirely be generalisable to other populations. However, given these
limitations, our study also has several strengths, including the longitudinal
design and the rigorous statistical techniques and methods applied, which are
specifically suited to investigating temporal relationships and to handling the
multivariate nature of the data. Lastly, our cohort contains highly detailed and
comprehensive variable measurements at each collection visit, and has both
concentration and mol% data for the fatty acids.

The role of total NEFA in the etiology of diabetes is well-documented.
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Paolisso1995a;@Salgin2012a;@Pankow2004a;@Charles1997a]. In a cross-sectional analysis of the
RISCK cohort, total NEFA had a negative association with insulin sensitivity and
a particularly strong negative association with beta cell function
[@Johns2014a]. Another cross-sectional study found a null association of
total NEFA on beta cell function [@Rebelos2015a], however they used C-peptide to
estimate beta cell function.
Experimentally, several potential mechanisms have been elucidated
for the role of NEFA on beta cell function, particularly for palmitic acid.
Prolonged exposure to elevated NEFA can induce apoptosis in the beta cells,
possibly through endoplasmic reticulum stress, formation of ceramides, and
generation of nitric oxide, as well as impairment of proinsulin production and
mitochondrial function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a]. 
One longitudinal studies also found an inverse association between total NEFA
and beta cell function over 10 years in 77 Japanese patients with type 2
diabetes [@Morita2012a], though they used C-peptide to estimate beta cell
function. The
present analysis is the first study to our knowledge to examine the longitudinal
association of a broad spectrum of individual NEFA on beta cell function in a
large cohort. We found that there was a strong signal of higher total NEFA,
palmitic acid, oleic acid, and linoleic acid modeled as concentrations with
lower beta cell function. No fatty acid as a mol% had associations with
beta cell function. Taken together, these results suggest that it is the
absolute size of the circulating NEFA fraction, irrespective of any specific 
composition of fatty acids, that is responsible for the hypothesized lipotoxic 
effects of NEFA on the beta cells.

@@

This study was supported by grants from the Canadian Diabetes Association (CDA),
the Canadian Institutes for Health Research, and the University of Toronto
Banting and Best Diabetes Centre; LWJ was supported by a CDA Doctoral Student
Research Award; RR was supported by a Heart and Stroke Foundation of Ontario
Mid-Career Investigator Award; SBH holds the CDA Chair in National Diabetes
Management and the Ian McWhinney Chair of Family Medicine Studies at the
University of Western Ontario; RPB holds a Tier II Canada Research Chair in
Brain Lipid Metabolism; AJH holds a Tier II Canada Research Chair in Diabetes
Epidemiology.

>>>>

Title: Association of non-esterified fatty acid composition with insulin
sensitivity and beta cell function in the Prospective Metabolism and Islet Cell
Evaluation (PROMISE) cohort

@@

@@

Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on the pathogenesis of diabetes, specifically on differences in insulin sensitivity and beta cell function over 6 years in a cohort of individuals who are at-risk for diabetes.

@@

At each examination, an 8-12 hour fasting blood sample was drawn from each
participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
minute and 2 hour blood draw. All blood samples were processed and frozen at
-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
Modular's enzymatic colorimetric tests (Roche Canada, Mississauga, Ontario, Canada).  Both insulin
and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
Basel, Switzerland) immunoassay analyser and electrochemiluminescence
immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
and the Des 31,32 circulating split form, and has a coefficient of variation (CV) of 9.3%.
Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
Diagnostics, Mississauga, Ontario, Canada) with a detection range of 0.11 (2 mg/dl) to 41.6 mmol/l.  The
inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
were categorised using the 2006 WHO criteria [@WHO2006].

@@

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables. Given the longitudinal design an
auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE
models, though other matrices (eg. exchangeable) had similar fit (data not
shown). GEE is well suited to longitudinal cohort studies given its capacity to
handle any missed visits and for the correlation of variables between time points. 
See the ESM Methods for more detail on GEE.
The predictor variables and continuous covariates were
scaled (mean centered and standardised). The NEFA variables were classified as 
*time-independent* (held constant) as they were measured only at the baseline 
visit, while the outcome variables were set as *time-dependent*. No imputation
was conducted on missing values and incident diabetes cases were excluded from 
the analysis (follow-up data were included until date of diagnosis). A GEE model
is applied for each combination of outcome and NEFA (as either mol% or nmol/ml).

@@

All analyses were performed using R `r getRversion()` [@Rbase], along with the R
packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a]
and the package pls `r packageVersion('pls')` for the PLS analysis.
The R code and output for all analyses of this manuscript is available at 
https://dx.doi.org/10.6084/m9.figshare.3472433. Results were considered
statistically significance at p<0.05, after adjusting for multiple testing using
the Benjamini-Hochberg False Discovery Rate (FDR) [@Benjamini1995a]. STROBE was used
as a guideline for reporting [@Vandenbroucke2007b].

@@

sig_fa_unadj <- gee_results %>% 
    dplyr::filter(model == "Unadjusted", p.value < 0.05) %>% 
    dplyr::select(Yterms, Xterms, unit) %>% 
    dplyr::mutate(Xterms = ifelse(Xterms == "Total", "total NEFA", as.character(Xterms)))
sig_fa_adj <- gee_results %>% 
    dplyr::filter(model == "Adjusted", p.value < 0.05) %>% 
    dplyr::select(Yterms, Xterms, unit) %>% 
    dplyr::mutate(Xterms = ifelse(Xterms == "Total", "total NEFA", as.character(Xterms)))

dm_each_visit <- calc_dm_each_visit()
```

@@

For this study, there were `r female$npct` females and `r european$npct` had
European-ancestry (see `r cite_t("basic")`). The mean follow-up time was 
`r calc_followup_time()` years, where `r calc_pct_full_visits()`% of 
participants attended all three visits.
The average age of the participants at baseline was `r aide::average(ds0yr$Age)`
years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
`r aide::average(ds0yr$BMI)` kg/m^2^ (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
participants, `r famhist$npct`, had a family history of diabetes. Insulin
sensitivity and beta cell function measures showed a significant median
decline from baseline to 6-years of between `r chg_outcomes$chg` 
(p`r chg_outcomes$p`, based on a time-adjusted GEE model) in this analysis.
Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
and `r calc_incid_dysgly("ConvertPreDM")` incident cases of diabetes and
pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
cases were excluded from GEE analyses (1, 28, and 13 cases were excluded at
0, 3, and 6 year visits, respectively).
Follow-up data were included until date of diagnosis.
Certain medications (e.g. lipid medications such as statins or
fibrates) could influence the association between NEFA and the outcomes.
There were `r meds_use[1]` and `r meds_use[2]` participants who used lipid 
modifying medications at the baseline and 6-year visit,
respectively. Use of these medications did not influence NEFA composition, the
outcomes, or their associations as analysed from the GEE models (data not shown).

@@

We used indices of insulin sensitivity and beta cell function estimated from
glucose and insulin values during OGTTs. While these indices have all been
validated against gold standard measures, it is possible our measures may
contain some degree of measurement misclassification and are thus not sensitive
enough to identify potential associations with NEFA.
Finally, our cohort consists of individuals at-risk for diabetes,
who are primarily female and European-ancestry and, while sex and ethnicity did
not influence the results presented from this cohort, our findings may not
entirely be generalisable to other populations. However, given these
limitations, our study also has several strengths, including the longitudinal
design and the rigorous statistical techniques and methods applied, which are
specifically suited to investigating temporal relationships and to handling the
multivariate nature of the data. Lastly, our cohort contains highly detailed and
comprehensive variable measurements at each collection visit, and has both
concentration and mol% data for the fatty acids.

The role of total NEFA in the etiology of diabetes is well-documented.
Epidemiological studies have shown that higher NEFA associate with lower insulin
secretion and a higher risk for developing diabetes 
[@Paolisso1995a;@Salgin2012a;@Pankow2004a;@Charles1997a]. In a cross-sectional analysis of the
RISCK cohort, total NEFA had a negative association with insulin sensitivity and
a particularly strong negative association with beta cell function
[@Johns2014a]. Another cross-sectional study found a null association of
total NEFA on beta cell function [@Rebelos2015a], however C-peptide was used to
estimate beta cell function in this study.
Experimentally, several potential mechanisms have been elucidated
for the role of NEFA on beta cell function, particularly for palmitic acid.
Prolonged exposure to elevated NEFA can induce apoptosis in the beta cells,
possibly through endoplasmic reticulum stress, formation of ceramides, and
generation of nitric oxide, as well as impairment of proinsulin production and
mitochondrial function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a]. 
One longitudinal study also found an inverse association between total NEFA
and beta cell function over 10 years in 77 Japanese patients with type 2
diabetes [@Morita2012a]; in this study C-peptide was used to estimate beta cell
function. The
present analysis is the first study to our knowledge to examine the longitudinal
association of a broad spectrum of individual NEFA on beta cell function in a
large cohort. We found that there was a strong signal of higher total NEFA,
palmitic acid, oleic acid, and linoleic acid modeled as concentrations with
lower beta cell function. No fatty acid as a mol% had associations with
beta cell function. Taken together, these results suggest that it is the
absolute size of the circulating NEFA fraction, irrespective of any specific 
composition of fatty acids, that is responsible for the hypothesized lipotoxic 
effects of NEFA on the beta cells.

@@

This study was supported by grants from the Canadian Diabetes Association (CDA;
Grant number: OG-3-14-4574-AH), the Canadian Institutes for Health Research
(Grant number: MOP-130458), and the University of Toronto
Banting and Best Diabetes Centre; LWJ was supported by a CDA Doctoral Student
Research Award; RR was supported by a Heart and Stroke Foundation of Ontario
Mid-Career Investigator Award; SBH holds the CDA Chair in National Diabetes
Management and the Ian McWhinney Chair of Family Medicine Studies at the
University of Western Ontario; RPB holds a Tier II Canada Research Chair in
Brain Lipid Metabolism; AJH holds a Tier II Canada Research Chair in Diabetes
Epidemiology.

## Data availability

The cohort datasets analysed during the current study are available from the
corresponding author on request.
